Star-galaxy Classification Using Deep Convolutional Neural Networks

### 2016-08-13 paper/main.tex
update paper

%\begin{figure}
%  \centering
%  \includegraphics[width=\columnwidth]{figures/sdss_color_color.pdf}
%  \caption{
%    Hexbin plots showing the color-color space distribution of galaxies (blue, top row) and stars (green, bottom row)
%    in the SDSS data set used in this analysis.
%    It is easy to see that there is significant overlap between stars
%    and galaxies in these color-color spaces. 
%  }
%  \label{fig:sdss_color_color}
%\end{figure}

@@

To create training images,
we obtain the image FITS files for SDSS fields containing these objects
in five photometric bands: $u$, $g$, $r$, $i$, and $z$.
We use the astrometry information in the FITS headers
in the \texttt{Montage}\footnote{http://montage.ipac.caltech.edu/} software
to align each image to the reference ($r$-band) image.
We then use \texttt{SExtractor} to find the pixel positions of
the 65,000 objects we have selected,
and to center each object in the cutout image.
{
Magnitudes in the SDSS photometric catalog are expressed as inverse hyperbolic
sine magnitudes~\citep[also known as luptitudes;][]{lupton1999modified},
and we follow the SDSS convention and convert all flux values to luptitudes.
}
Finally, in order to account for the effect of Galactic dust,
extinction corrections in magnitudes are applied
following \cite{schlegel1998maps}.
In the end, we have cutout images of size $48\times48$ pixels
with luptitude values in each pixel.
We note that we have experimented with increasing the pixel dimensions to
$60\times60$ and $72\times72$ pixels, but do not find noticeable
improvement in the performance of our model.

@@

%\begin{figure}
%  \centering
%  \includegraphics[width=\columnwidth]{figures/clens_color_color.pdf}
%  \caption{
%    Hexbin plots showing the color-color space distribution of
%    galaxies (blue, top row) and stars (green, bottom row)
%    in the CFHTLenS data set used in this analysis.
%  }
%  \label{fig:clens_color_color}
%\end{figure}

@@

\changed{
In this section, we first provide a brief overview of deep neural networks,
including convolutional neural networks (ConvNets).
We then describe the precise network architecture and the details of key
hyperparamters.
}

@@

Many deep learning models use feedforward neural network architectures with multiple layers,
where each neuron in one layer is connected to the neurons of the subsequent layer.
A schematic representation is shown in Figure~\ref{fig:neuron_b}.
All layers except the input and output layers are conveniently called hidden layers.
\sout{ With multiple hidden layers, deep neural networks can model complex nonlinear relationships, and they have had great success in multimedia tasks where humans traditionally outperformed computers, such as speech recognition, language processing, and image classification.  }

The goal of learning is to find a set of weights and biases such that,
given $N$ samples, the output from the network
$\bmath{y}=\left(y_1, y_2, \dots, y_N \right)$ approximates the desired output
$\hat{\bmath{y}}=\left(\hat{y}_1, \hat{y}_2, \dots, \hat{y}_N \right)$
as closely as possible for all input
$\bmath{X}=\left(\bmath{x}_1,\bmath{x}_2,\dots,\bmath{x}_N\right)$.
We can formulate this as the minimization of a loss function
$L(\bmath{y},\hat{\bmath{y}})$
over the training data.
In this work, we use \textit{cross-entropy}
\changed{
\citep[also called log loss;][]{murphy2012machine} as the loss function.
For binary classification, the cross-entropy per sample is given by
\begin{equation}
  L(y_j, \hat{y}_j) = -\hat{y}_j \log_2 y_j - (1 - \hat{y}_j) \log_2(1 - y_j),
  \label{eq:cross_entropy}
\end{equation}
}
where $\hat{y}_j$ is the actual truth value (\eg 0 or 1) of the $j$-th data, and
$y_j$ is the probability prediction made by the model.
We compute the loss function by taking the average of all cross-entropies in the sample.
Thus, the loss function becomes
\begin{equation}
\changed{
  L(\bmath{y}, \hat{\bmath{y}})
}
= - \frac{1}{N} \sum_{j=1}^{N} \hat{y}_j  \log_2 y_j
    + (1 - \hat{y}_j)  \log_2 (1 - \hat y_j).
  \label{eq:cross_entropy_all}
\end{equation}

@@

\changed{
Typically, a
}
pooling layer computes the maximum of a local $2\times2$ patch in a feature map
\changed{
\citep{krizhevsky2012imagenet}.
}
Since the pooling layer aggregates the activations of neighboring units from the previous layer,
it reduces the dimensionality of the feature maps and
makes the model invariant to small shifts and distortions.
Two or more layers of convolution and pooling are stacked,
followed by more convolutional and fully-connected layers.

@@

To make final classifications,
we use our ConvNet model to make 64 sets of predictions for
64 transformations of the input images:
4 rotations, 4 horizontal translations, and 4 vertical translations
(with random horizontal reflections).
Although we use an identical network architecture for all transformations,
we consider each set of predictions as separate results from different models.
Finally, we use a model combination technique known as
Bayesian Model Combination~\citep[BMC;][]{Monteith2011}, which
uses Bayesian principles to generate an ensemble combination of different models.
Although the data augmentation step in Section~\ref{sec:data_augmentation} should
make our ConvNet model invariant to these types of transformations,
we find that applying BMC still results in a significant increase in performance.
\changed{
For a thorough discussion of BMC, we refer the reader to \citet{Monteith2011}.
}

@@

\sout{ TPC is a part of a publicly available software package called \textsc{MLZ}%\footnote{http://pythonhosted.org/MLZ/} (Machine Learning for Photo-$z$).
}

@@

\changed{
Probabilistic classifiers, rather than only assigning discrete labels to each
source, produce a continuous probability distribution of whether each source is
a star or a galaxy.
To evaluate the performance of probabilistic classifiers,
many studies~\citep[\eg][]{henrion2011bayesian, Fadely2012} convert
probability estimates into class labels by choosing a probability
threshold, \eg a source is classified as a star if $P_{\rmn{class}} < 0.5$,
and a galaxy if $P_{\rmn{class}} > 0.5$.
However, using a fixed threshold ignores the model's operating conditions,
such as science requirements, misclassification costs, and stellar distribution.
Furthermore, the probability threshold of 0.5 is not necessarily optimal for
an unbalanced data set, where galaxies outnumber stars.
Following \citet{kim2015hybrid}, we use performance metrics that are well-suited
for probabilistic classifiers: Area Under the Curve (AUC)
for the Receiver Operating Characteristic (ROC) curve, completeness and purity,
and the Mean Squared Error (MSE).
To evaluate 
we also use the calibration error and the absolute error in the estimation
of number of galaxies as performance metrics.
}

@@

@@

\begin{table*}
  \caption{
    A summary of the classification performance metrics
    as applied to the CFHTLenS data.
    The definition of the metrics is summarized in Table~\ref{table:metrics}.
    The bold entries highlight the best performance values within each column.
  }
  \centering
  \begin{tabular}{l c c c c c c c}
    \hline
    classifier & AUC & MSE & $p_{g}(c_g=0.96)$ & $c_{s}(p_s=0.97)$ & CAL & $ |\Delta N_g|/N_g$ &
\changed{
log loss
}
\\
    \hline
    ConvNet                       & \textbf{0.9948} & 0.0112          & \textbf{0.9972} & 0.8971          & \textbf{0.0197}  & \textbf{0.0029} &
\changed{
\textbf{0.0441}
}
\\
    TPC${}_{\rmn{morph}}$         & 0.9924          & \textbf{0.0109} & 0.9963          & \textbf{0.9268} & 0.0245           & 0.0056 &
\changed{
0.0809
}
\\
    TPC${}_{\rmn{phot}}$          & 0.9876          & 0.0189          & 0.9927          & 0.8044          & 0.0266           & 0.0101 &
\changed{
0.1085
}
\\
    \hline
  \end{tabular}
  \label{table:clens_metrics}
\end{table*}

@@

\changed{
Note that $p_{g}(c_g=0.96)$ and $c_{s}(p_s=0.97)$ require adjusting
threshold values (\ie probability cuts), while other metrics do not.
To obtain a galaxy completeness of $c_g=0.96$, we choose the threshold values
0.9972, 0.9963, and 0.9927 for ConvNet, TPC${}_{\rmn{morph}}$, TPC${}_{\rmn{phot}}$,
respectively;
for star purity $p_s=0.97$, we choose 0.6990, 0.5297, and 0.8570 for
ConvNet, TPC${}_{\rmn{morph}}$, TPC${}_{\rmn{phot}}$, respectively.
In Table ? and ?, we present example.
}

@@

\begin{table*}
  \caption{
    A summary of the classification performance metrics
    as applied to the SDSS data.
  }
  \centering
  \begin{tabular}{l c c c c c c c}
    \hline
    classifier & AUC & MSE & $p_{g}(c_g=0.96) $ & $c_{s}(p_s=0.97)$ & CAL & $ |\Delta N_g|/N_g$ & \changed{log loss} \\
    \hline
    ConvNet               & 0.9952          & 0.0182          & 0.9915          & 0.9500          & \textbf{0.0243} & 0.0157 & \changed{\textbf{0.0731}} \\
    TPC${}_{\rmn{morph}}$ & \textbf{0.9967} & \textbf{0.0099} & \textbf{0.9977} & \textbf{0.9810} & 0.0254          & \textbf{0.0044} & \changed{0.0914} \\
    TPC${}_{\rmn{phot}}$  & 0.9886          & 0.0283          & 0.9819          & 0.8879          & 0.0316          & 0.0160 & \changed{0.1372} \\
    \hline
  \end{tabular}
  \label{table:sdss_metrics}
\end{table*}

@@

@@

\changed{
This work has made use of the Theano library (http://deeplearning.net/software/theano/),
the Lasagne library (http://lasagne.readthedocs.io/),
and the Nolearn library (http://pythonhosted.org/nolearn/).
}

>>>>

@@

To create training images,
we obtain the image FITS files for SDSS fields containing these objects
in five photometric bands: $u$, $g$, $r$, $i$, and $z$.
We use the astrometry information in the FITS headers
in the \texttt{Montage}\footnote{http://montage.ipac.caltech.edu/} software
to align each image to the reference ($r$-band) image.
We then use \texttt{SExtractor} to find the pixel positions of
the 65,000 objects we have selected,
and to center each object in the cutout image.
\changed{
Magnitudes in the SDSS photometric catalog are expressed as inverse hyperbolic
sine magnitudes~\citep[also known as luptitudes;][]{lupton1999modified},
and we follow the SDSS convention and convert all flux values to luptitudes.
}
Finally, in order to account for the effect of Galactic dust,
extinction corrections in magnitudes are applied
following \cite{schlegel1998maps}.
In the end, we have cutout images of size $48\times48$ pixels
with luptitude values in each pixel.
We note that we have experimented with increasing the pixel dimensions to
$60\times60$ and $72\times72$ pixels, but do not find noticeable
improvement in the performance of our model.

@@

@@

\changed{
In this section, we first provide a brief overview of deep neural networks
including convolutional neural networks.
We then describe the precise network architecture and the details of key
hyperparamters.
}

@@

Many deep learning models use feedforward neural network architectures with multiple layers,
where each neuron in one layer is connected to the neurons of the subsequent
layer~\citep{lecun2015deep}.
A schematic representation is shown in Figure~\ref{fig:neuron_b}.
All layers except the input and output layers are conveniently called hidden layers.
\sout{ With multiple hidden layers, deep neural networks can model complex nonlinear relationships, and they have had great success in multimedia tasks where humans traditionally outperformed computers, such as speech recognition, language processing, and image classification.  }

\changed{
We
}
find a set of weights and biases such that,
given $N$ samples, the output from the network
$\bmath{y}=\left(y_1, y_2, \dots, y_N \right)$ approximates the desired output
$\hat{\bmath{y}}=\left(\hat{y}_1, \hat{y}_2, \dots, \hat{y}_N \right)$
as closely as possible for all input
$\bmath{X}=\left(\bmath{x}_1,\bmath{x}_2,\dots,\bmath{x}_N\right)$.
We can formulate this as the minimization of a loss function
$L(\bmath{y},\hat{\bmath{y}})$
over the training data.
In this work, we use \textit{cross-entropy}
\changed{
\citep[also called log loss;][]{murphy2012machine} as the loss function.
For binary classification, the cross-entropy per sample is given by
\begin{equation}
  L(y_j, \hat{y}_j) = -\hat{y}_j \log_2 y_j - (1 - \hat{y}_j) \log_2(1 - y_j),
  \label{eq:cross_entropy}
\end{equation}
}
where $\hat{y}_j$ is the actual truth value (\eg 0 or 1) of the $j$-th data, and
$y_j$ is the probability prediction made by the model.
We compute the loss function by taking the average of all cross-entropies in the sample.
Thus, the loss function becomes
\begin{equation}
\changed{
  L(\bmath{y}, \hat{\bmath{y}})
}
= - \frac{1}{N} \sum_{j=1}^{N} \hat{y}_j  \log_2 y_j
    + (1 - \hat{y}_j)  \log_2 (1 - \hat y_j).
  \label{eq:cross_entropy_all}
\end{equation}

@@

\changed{
Typically, a
}
pooling layer computes the maximum of a local $2\times2$ patch in a feature map
\changed{
\citep{krizhevsky2012imagenet}.
}
Since the pooling layer aggregates the activations of neighboring units from the previous layer,
it reduces the dimensionality of the feature maps and
makes the model invariant to small shifts and distortions
\changed{
\citep{boureau2010theoretical}
}
.
Two or more layers of convolution and pooling are stacked,
followed by more convolutional and fully-connected layers.

@@

To make final classifications,
we use our ConvNet model to make 64 sets of predictions for
64 transformations of the input images:
4 rotations, 4 horizontal translations, and 4 vertical translations
(with random horizontal reflections).
Although we use an identical network architecture for all transformations,
we consider each set of predictions as separate results from different models.
Finally, we use a model combination technique known as
Bayesian Model Combination~\citep[BMC;][]{Monteith2011}, which
uses Bayesian principles to generate an ensemble combination of different models.
Although the data augmentation step in Section~\ref{sec:data_augmentation} should
make our ConvNet model invariant to these types of transformations,
we find that applying BMC still results in a significant increase in performance.
\changed{
For a thorough discussion of BMC, we refer the reader to \citet{Monteith2011}
(See also \citet{carrascokind2014exhausting} for application of BMC to
photometric estimation, and \citet{kim2015hybrid} for combining star-galaxy
classifiers).
}

@@

@@

\changed{
Probabilistic classifiers, rather than only assigning discrete labels to each
source, produce a continuous probability distribution of whether each source is
a star or a galaxy.
To evaluate the performance of probabilistic classifiers,
many studies~\citep[\eg][]{henrion2011bayesian, Fadely2012} convert
probability estimates into class labels by choosing a probability
threshold, \eg a source is classified as a star if $P_{\rmn{class}} < 0.5$,
and a galaxy if $P_{\rmn{class}} > 0.5$.
However, using a fixed threshold ignores the model's operating conditions,
such as science requirements, misclassification costs, and stellar distribution.
Furthermore, the probability threshold of 0.5 is not necessarily optimal for
an unbalanced data set, where galaxies outnumber stars.
}

\changed{
Following \citet{kim2015hybrid}, we use performance metrics that are well-suited
for probabilistic classifiers: Area Under the Curve (AUC)
for the Receiver Operating Characteristic (ROC) curve, completeness and purity,
and the Mean Squared Error (MSE).
A good probabilistic classifier should also provide well-calibrated
posterior probabilities.
Thus, to evaluate calibration performance,
we also use the calibration error and the absolute error in the estimation
of number of galaxies as performance metrics.
}

@@

\changed{
Let $N_g$ be the number of true galaxies classified as galaxies,
and $M_g$ the number of true galaxies classified as stars.
Then the galaxy \textit{completeness} $c_g$ (also called recall or sensitivity)
is given by
}
\begin{equation}
c_g = \frac{N_g}{N_g + M_g}.
\end{equation}
\changed{
Let $M_s$ be the number of true stars classified as galaxies.
Then the galaxy \textit{purity} $p_g$ (also called precision or positive
predictive value) is given by
}
\begin{equation}
p_g = \frac{N_g}{N_g + M_s}.
\end{equation}
\changed{
We define the star completeness and purity in a similar way.
As discussed in our previous work~\citep{kim2015hybrid},
we adopt weak lensing science requirements of the DES~\citep{soumagnac2013star},
and compute $p_g$ at $c_g=0.960$ and $c_s$ at $p_s=0.970$.
}

@@

\begin{table*}
  \caption{
    A summary of the classification performance metrics
    as applied to the CFHTLenS data.
    The definition of the metrics is summarized in Table~\ref{table:metrics}.
    The bold entries highlight the best performance values within each column.
\changed{
    Note that $p_{g}(c_g=0.96)$ and $c_{s}(p_s=0.97)$ require adjusting
    threshold values (\ie probability cuts), while other metrics do not.
    To obtain a galaxy completeness of $c_g=0.96$, we choose the threshold values
    0.9972, 0.9963, and 0.9927 for ConvNet, TPC${}_{\rmn{morph}}$, TPC${}_{\rmn{phot}}$,
    respectively;
    for star purity $p_s=0.97$, we choose 0.6990, 0.5297, and 0.8570 for
    ConvNet, TPC${}_{\rmn{morph}}$, TPC${}_{\rmn{phot}}$, respectively.
}
  }
  \centering
  \begin{tabular}{l c c c c c c c}
    \hline
    classifier & AUC & MSE & $p_{g}(c_g=0.96)$ & $c_{s}(p_s=0.97)$ & CAL & $ |\Delta N_g|/N_g$ &
\changed{
log loss
}
\\
    \hline
    ConvNet                       & \textbf{0.9948} & 0.0112          & \textbf{0.9972} & 0.8971          & \textbf{0.0197}  & \textbf{0.0029} &
\changed{
\textbf{0.0441}
}
\\
    TPC${}_{\rmn{morph}}$         & 0.9924          & \textbf{0.0109} & 0.9963          & \textbf{0.9268} & 0.0245           & 0.0056 &
\changed{
0.0809
}
\\
    TPC${}_{\rmn{phot}}$          & 0.9876          & 0.0189          & 0.9927          & 0.8044          & 0.0266           & 0.0101 &
\changed{
0.1085
}
\\
    \hline
  \end{tabular}
  \label{table:clens_metrics}
\end{table*}

@@

@@

\begin{table*}
  \caption{
    A summary of the classification performance metrics
    as applied to the SDSS data.
\changed{
    To obtain a galaxy completeness of $c_g=0.96$, we choose the threshold values
    0.7558, 0.9989, and 0.9360 for ConvNet, TPC${}_{\rmn{morph}}$, TPC${}_{\rmn{phot}}$,
    respectively;
    for star purity $p_s=0.97$, we choose 0.6046, 0.0547, and 0.7449 for
    ConvNet, TPC${}_{\rmn{morph}}$, TPC${}_{\rmn{phot}}$, respectively.
}
  }
  \centering
  \begin{tabular}{l c c c c c c c}
    \hline
    classifier & AUC & MSE & $p_{g}(c_g=0.96) $ & $c_{s}(p_s=0.97)$ & CAL & $ |\Delta N_g|/N_g$ & \changed{log loss} \\
    \hline
    ConvNet               & 0.9952          & 0.0182          & 0.9915          & 0.9500          & \textbf{0.0243} & 0.0157 & \changed{\textbf{0.0731}} \\
    TPC${}_{\rmn{morph}}$ & \textbf{0.9967} & \textbf{0.0099} & \textbf{0.9977} & \textbf{0.9810} & 0.0254          & \textbf{0.0044} & \changed{0.0914} \\
    TPC${}_{\rmn{phot}}$  & 0.9886          & 0.0283          & 0.9819          & 0.8879          & 0.0316          & 0.0160 & \changed{0.1372} \\
    \hline
  \end{tabular}
  \label{table:sdss_metrics}
\end{table*}

@@

\changed{
Furthermore, our ConvNet is a supervised algorithm; and one of the criticisms
of supervised techniques is their difficulty in extrapolating past the limits
of available spectroscopic training data.
Since it is difficult to assess the classification performance without a deeper
spectroscopic sample, we have evaluated the performance on a test set
that is drawn from the same underlying distribution as the spectroscopic sample.
However, when our ConvNet model
---trained on sources from a spectroscopic sample---
is applied to a photometric sample
---which is often fainter than our training set--- 
the performance of ConvNet will be less reliable.
Combining our ConvNet model with unsupervised methods (\eg a template fitting
method) in this meta-classification framework will help improve the efficacy of
star-galaxy classification beyond the limits of spectroscopic training data.
}

@@

\changed{
This work has made use of the Theano library
(\href{http://deeplearning.net/software/theano/}{http://deeplearning.net/software/theano/}),
the Lasagne library
(\href{http://lasagne.readthedocs.io/)}{http://lasagne.readthedocs.io/}),
and the Nolearn library
(\href{http://pythonhosted.org/nolearn/}{http://pythonhosted.org/nolearn/}).
}
